---
title: "Pharmacokinetic exposure by ethnicity"
description: |
  The importance of understanding the scientific context.
author:
  - name: Marc Vandemeulebroecke, Mark Baillie, Alison Margolskee and Baldur Magnusson  
    url: https://graphicsprinciples.github.io/
date: "`r Sys.Date()`"
draft: true
output:
  radix::radix_article:
    self_contained: false
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "##",
  R.options = list(width = 60)
)

## Specify libraries and ggplot theme
library(radix)
library(tidyverse)
library(RColorBrewer)
library(caTools)
library(gridExtra)
library(data.table)
library(grid)

source("../../R/paper_theme.R")
```


This case study is inspired by a publication comparing the pharmacokinetic exposure of a drug in Caucasian vs. Japanese subjects. The same single oral dose was administered to all subjects. The original graph displays mean +/- standard deviations (SD) of the plasma concentration over time, grouped by ethnicity (Caucasian or Japanese).  

After reading in and performing a number of small number of data transformations we recreate the original graph in Figure 1. 


```{r, include = TRUE}
## Read in data and define order for factors
my_data <- read_csv("../../data/402_case2_PKdataset.csv") %>%
  filter(CMT == 2 & DOSE == 300 & PART == 1 & STUDY == 1) %>%
  mutate(
    TRTACT = factor(TRTACT, levels = rev(unique(TRTACT[order(DOSE)]))),
    ETHN = factor(ETHN),
    ETHN = factor(ETHN, levels = rev(levels(ETHN)))
  ) 
```

```{r, fig.cap="Pharmacokinetic exposure by ethnicity."}

## Plot mean and error bars (SD) on a linear scale
my_data %>%
  ggplot(aes(x = NOMTIME, y = LIDV, group = interaction(ETHN, CYCLE))) + 
  stat_summary(geom = "errorbar", width = 2,  
               fun.data = mean_sdl, fun.args = list(mult = 1)) +
  stat_summary(geom = "line", size = 0.5, 
               fun.y = mean) + 
  stat_summary(aes(fill = ETHN), geom = "point", size = 1.5, 
               fun.y = mean, stroke = 0.5, shape = 21) + 
  scale_fill_manual(values = c("white", "black")) + 
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24, 36, 48, 72)) + 
  xlab("Time (h)") + 
  ylab("Concentration (ng/mL)\nMean (SD)") + 
  theme_bw(base_size = 8) + 
  theme(legend.title = element_blank(), 
        legend.position = "bottom",
        legend.box.spacing = unit(0, "mm")) 
```



This time, let us start with Law 1. What is the purpose of this graph? For drugs that are mainly developed in a Caucasian population, Japanese drug regulation requires sponsors to investigate whether the pharmacokinetics (PK) are similar or different between Caucasian and Japanese populations. The purpose of this graph is to help address this question. Looking at it, we may be tempted so say that the PK are reasonably similar. But are they really? If they are not, then in what way?

This leads us to Law 2: show the data clearly. The graphical attributes in Figure 8A appear to be wisely chosen: the symbols and labels are clear, the gridlines are supportive and stay in the background, and there is no unnecessary adornment. However, at least two things obscure the answer to our question of interest. First, the concentrations are plotted on a linear scale, which makes it difficult to distinguish them at the lower end of the profile. Concentrations should be plotted on a logarithmic scale because they are log-normally distributed. In fact, for concentrations in particular (but not generally for any log-normally distributed data) we should produce both types of display: one on a log-linear scale (to assess the elimination phase) and one on a linear scale (to see the peak more clearly). Second, it is hard to determine whether any differences are significant when standard deviations are plotted instead of standard errors or confidence intervals. Standard deviations show the variation in the data; they do not shrink when more data is collected. Standard errors show the variation in the means. Confidence intervals may be the best choice as they directly show the uncertainty about the means.



```{r, fig.cap="Figure 1 looks fair at first sight but reveals important information after two simple changes, namely, scaling the y-axis differently and plotting confidence intervals instead of standard deviations"}

## Mean (95% CI) error bars, log scale
my_data %>%
  ggplot(aes(x = NOMTIME, y = LIDV, group = interaction(ETHN, CYCLE))) + 
  stat_summary(geom = "errorbar", width = 2,  
               fun.data = mean_cl_normal, fun.args = list(mult = 1)) +
  stat_summary(geom = "line", size = 0.5, 
               fun.y = mean) + 
  stat_summary(aes(fill = ETHN), geom = "point", size = 1.5, 
               fun.y = mean, stroke = 0.5, shape = 21) + 
  scale_fill_manual(values = c("white", "black")) + 
  scale_x_continuous(breaks = c(0, 4, 8, 12, 24, 36, 48, 72)) + 
  xlab("Time (h)") + ylab("Concentration (ng/mL)\nMean (95% CI)") + 
  guides(color = guide_legend(title = "Dose")) + 
  scale_y_log10() + 
  annotation_logticks(base = 10, sides = "l", color = rgb(0.5, 0.5, 0.5)) +
  theme_bw(base_size = 8) + 
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.box.spacing = unit(0, "mm")) 


```




These issues have been fixed in Figure 8B (we only show the log-linear version). To reduce cluttering, the ticks at the end of the whiskers have also been omitted (non-data ink). If the graph displayed more than two profiles, we might consider replacing the whiskers by (shaded) confidence bands, or separating the graph out in panels or “small multiples” (see bottom of the backside of the Cheat Sheet on https://graphicsprinciples.github.io). From Figure 8B it appears that the higher concentrations are not meaningfully different, but the elimination phase does differ between the two ethnicities. This could also translate to different average exposures. That is, based on two simple changes in the plot, we now see answers emerging for our initial question about PK differences. We see them emerging with respect to three key PK characteristics: peak, elimination/trough, and overall exposure. While (depending on the drug) similarity in the peak may be reassuring from a safety point of view, a lower overall exposure in Japanese subjects could be a concern for efficacy.

Moving on to Law 3, let us now make the message obvious. We could choose a completely different graph type to hone in on the message. Figure 8C shows the three (non-compartmentally derived) quantities Cmax, Ctrough and AUClast with 95% confidence intervals by ethnicity. Clearly, Ctrough and AUC are different between the two ethnicities, but Cmax is not. We could derive these quantities from a compartmental model fit and produce the same plot. Or we could go one step further and show directly their geometric mean ratio, Japanese vs. Caucasian subjects, as in Figure 8D. This last plot answers the initial question most succinctly, and its graphical appearance has also been further simplified (no frame, minimal gridlines, mildly highlighted line of equality), to not distract from the message. In practice, Figures 8B, 8C and/or 8D together may be most informative, covering the time course as well as differences in key parameters.



```{r, fig.cap="Key messages are made more obvious by directly plotting the pharmacokinetic parameters of interest."}

theme_set(theme_bw(base_size = 10))

## Plot cmax, ctrough, AUClast dots (95% CI) in separate panels

Cmax <- my_data %>% 
  filter(!is.na(LIDV))  %>%
  group_by(ID, ETHN) %>%
  summarize(Cmax = max(LIDV),
            Ctrough = min(LIDV))

Ctrough <- my_data %>% 
  filter(!is.na(LIDV))  %>%
  group_by(ID, ETHN) %>%
  summarize(Ctrough = min(LIDV))

AUClast <- my_data %>%
  filter(!is.na(LIDV)) 

AUClast <- data.frame(stack(sapply(split(AUClast, AUClast$ID), function(df) trapz(df$TIME, df$LIDV))))
names(AUClast) <- c("AUC", "ID")

AUClast$ID <- as.numeric(as.character(AUClast$ID))
AUClast <- AUClast[order(AUClast$ID), ]
AUClast <- merge(AUClast, unique(my_data[c("ID", "ETHN")]), by = "ID")


gg1 <- Cmax %>% 
  ggplot(aes(x = ETHN, y = Cmax, group = ETHN)) + 
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0) + 
  stat_summary(shape = 21, fill = "white", size = 0.2) + 
  labs(title = "Cmax (ng/mL)", subtitle = "Mean (95% CI)") + 
  scale_y_log10(breaks = c(0.3, 1, 3, 10, 30, 100, 300, 1000, 3000), limits = c(60,300)) +   annotation_logticks(base = 10, sides = "l", color = rgb(0.5, 0.5, 0.5)) +
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank(),
        plot.subtitle = element_text(color = rgb(0.3, 0.3, 0.3)))

gg2 <- gg1 %+% #Cmax %+% 
  aes(x = ETHN, y = Ctrough) + 
  labs(title = "Ctrough (ng/mL)") + 
  scale_y_log10(breaks = c(0.3, 1, 3, 10, 30, 100, 300, 1000, 3000), limits = c(0.6, 9)) 


gg3 <- gg1 %+% AUClast %+% aes(x = ETHN, y = AUC) + 
  labs(title = "AUClast (h.(ng/mL))") + 
  scale_y_log10(breaks = c(0.3, 1, 3, 10, 30, 100, 300, 1000, 3000), limits = c(500, 1500)) 

grid.arrange(arrangeGrob(gg1, gg2, gg3, nrow = 1), nrow = 1)

```

```{r, fig.cap="Or even only their ratios."}
##############################################################################
## Forest plot of Ratios of Japanese:Caucasian for Cmax, Ctrough, AUClast
##############################################################################
Cmax2  <- Cmax %>% 
  mutate(DV = Cmax, logDV = log(Cmax), LABEL = "Cmax") %>% 
  select(c("ID", "ETHN", "DV", "logDV", "LABEL"))

glimpse(Cmax2)

results <- t.test(logDV ~ ETHN, Cmax2)

PKmetrics <- data.frame( y2.5 = exp(results$conf.int[1]), 
                         y97.5 = exp(results$conf.int[2]), 
                         ymean = exp(as.numeric(results$estimate[1] - results$estimate[2])), 
                         var = "Cmax",
                         unit = "ng/mL")

Ctrough2 <- Ctrough %>% 
  mutate(DV = Ctrough, logDV = log(Ctrough), LABEL = "Ctrough") %>% 
  select(c("ID", "ETHN", "DV", "logDV", "LABEL"))

results <- t.test(logDV ~ ETHN, Ctrough2)

PKmetrics <- rbind(PKmetrics, 
                   data.frame( y2.5 = exp(results$conf.int[1]), 
                               y97.5 = exp(results$conf.int[2]), 
                               ymean = exp(as.numeric(results$estimate[1] - results$estimate[2])), 
                               var = "Ctrough",
                               unit = "ng/mL"))

AUClast2 <- AUClast %>% 
  mutate(DV = AUC, logDV = log(AUC), LABEL = "AUClast") %>% 
  select(c("ID", "ETHN", "DV", "logDV", "LABEL"))

results <- t.test(logDV ~ ETHN, AUClast2)

PKmetrics <- rbind(PKmetrics, 
                   data.frame( y2.5 = exp(results$conf.int[1]), 
                               y97.5 = exp(results$conf.int[2]), 
                               ymean = exp(as.numeric(results$estimate[1] - results$estimate[2])), 
                               var = "AUClast",
                               unit = "h.ng/mL"))

PKmetrics$var <- factor(PKmetrics$var, levels = c("AUClast", "Ctrough", "Cmax"))

PKmetrics %>%
  ggplot(aes(x = var, y = ymean, ymin = y2.5, ymax = y97.5)) + 
  paper_theme() +
  geom_hline(yintercept = 1, size = 1, colour = "red", alpha = 0.1) + 
  scale_y_log10(breaks = c(0.25, 0.5, 1, 2, 4)) +
  geom_point() + 
  geom_errorbar(width = 0) + 
  xlab("") +
  scale_x_discrete(breaks = NULL, labels = NULL) + 
  ylab("Ratios between Japanese and Caucasian Mean (95% CI)") + 
  geom_text(aes(x = var, y = 0.175, label = paste0(var, "\n(", unit, ")" ) )) +
  theme(
        panel.grid.major.x = element_line(color = "gray", size = 0.2), 
        panel.grid.major.y = element_blank(), 
        panel.border = element_blank(),
        axis.title.x = element_text(size = 11, vjust = -1.25),
        axis.title.y = element_blank()
        ) + 
  coord_flip(ylim = c(0.15, 5)) 
```
